Re-examining cross-cultural similarity judgments using lexical co-occurrence

### 2023-01-31 code/4_paper/manuscript/manuscript.tex
manuscript edits

\begin{abstract}
Is ``cow'' more closely related to ``grass'' or ``chicken''? Speakers of
different languages judge similarity in this context differently, but
why? One possibility is that cultures co-varying with these languages
induce differences in conceptualizations of similarity. Specifically,
East Asian cultures may promote reasoning about thematic similarity, by
which cow and grass are more related, whereas Western cultures may bias
judgments toward taxonomic relations, like cow-chicken. This difference
in notions of similarity is the consensus interpretation for
cross-cultural variation in this paradigm. We consider, and provide
evidence for, an alternative possibility, by which notions of similarity
are similar across contexts, but the statistics of the environment vary.
On this account, similarity judgments are guided by co-occurrence in
experience, and observing or hearing about cows and grass or cows and
chickens more often could induce preferences for these groupings, and
account in part for apparent differences in notions of similarity across
contexts.

@@

\hypertarget{cross-cultural-variation-in-similarity}{%
\subsection{Cross-cultural variation in
similarity}\label{cross-cultural-variation-in-similarity}}

The way we partition our continuous experiences of the world into
categories, and which kind of similarity relationships we privilege to
organize such partitions, vary across cultures. One way this
cross-cultural difference has been observed is with a task comparing
preference for taxonomic and thematic similarity between East Asian and
Western participants.\footnote{Taxonomic categorization is based on the
  similarity of attributes, for example, similar perceptual properties,
  like shared color or shape, among objects.In contrast, thematic
  categorization is based on causal, spatial, and temporal relationships
  among objects (Markman \& Hutchinson, 1984).} Ji, Zhang, \& Nisbett
(2004) found that in a word triad task (choose two out of three words
that are more related to one another), Chinese participants preferred
thematic matching more compared to European Americans. In a pictorial
version of this task, Chiu (1972) found that Chinese children (9-10
years old) are also more likely to choose thematic matches compared to
their American counterparts. This cross-cultural difference is also
observed in novel object categorization, with Chinese participants
preferring to group by family resemblance across multiple features and
Americans preferring a single-feature rule (Norenzayan, Smith, Kim, \&
Nisbett, 2002).

Previous research has linked the observed difference in similarity
judgment to related work showing tendencies toward analytic processing
(emphasizing objects and their properties) in Western cultures and
holistic processing in East Asian cultures (emphasizing relationships
between objects and their context) (see also Nisbett (2003)).
Cross-cultural difference in similarity judgment is thus connected to
well-documented work showing East Asian participants exhibiting a higher
level of sensitivity to context than their Western counterparts when
reproducing drawings from memory (Ji, Peng, \& Nisbett, 2000); visually
exploring naturalistic scenes (Chua, Boland, \& Nisbett, 2005);
describing scenes (Masuda \& Nisbett, 2001); and in explaining the
causes of ambiguous behaviors (Choi, Nisbett, \& Norenzayan, 1999). This
view ascribes cross-cultural differences in similarity judgment to
variation in how each culture (or group of culture, i.e.~East Asian vs
Western) \textbf{conceptualize} similarity -- with East Asians
preferring thematic relations more than Westerners.

Alternatively, these judgments could be shaped by cross-cultural
differences in the \textbf{inputs} to similarity judgment, namely, the
statistics of the environment, and accordingly the content of everyday
experiences. Perhaps when faced with the triad task, participants from
all cultures follow the same process for conceptualizing similarity, but
rely on language or culture-specific inputs to this process. If we
observe a difference in categorization between East Asian and Western
participants, it could be that members of both groups use the same
procedure (considering similarity that is influenced by both taxonomic
and thematic relations), but the inputs to this procedure differ between
cultures, with East Asian participants exposed to more instances of
thematic similarity than their Western counterparts. We might also
expect that both conceptualization and inputs of similarity play a role
in driving cross-cultural differences -- East Asian participants may
both be exposed to more instances of thematic similarity, and prefer to
conceptualize similarity as thematic relations more so than Westerners.

@@

To test alternative hypotheses involving variation of inputs to
similarity judgment, we need a way to operationalize variation in
exposure to thematic vs taxonomic relations. While this metric is
difficult to directly measure, language statistics can provide a rough
proxy. Language statistics is useful in that it is part of the inputs to
everyday experiences (through spoken and written language), may afford
many of the `experiences' that people have with infrequently encountered
items, like cows or helicopters, and provides accessible measures.
Previous work also suggests that using language statistics (such as
lexical co-occurrence or cosine distance of word embeddings) as a proxy
can be a good predictor for similarity reasoning. Griffiths, Steyvers,
\& Tenenbaum (2007) showed that a model trained on word-document
co-occurrence can predict word association and the effects of semantic
association on a variety of linguistic tasks. Lexical semantic models
that are constructed using lexical co-occurrence (in comparison to
annotated relations) have been shown to perform well on predicting human
judgments about similarity between word pairs that are thematically or
taxonomically related (Rohde, Gonnerman, \& Plaut, 2006). Word
embeddings like fastText have been demonstrated to be good predictors
for similarity judgments (Liu, Feng, Wu, Chan, \& Fulton (2019),
Jatnika, Bijaksana, \& Suryani (2019)).

@@

Our study tests the hypothesis that, rather than (or in addition to) the
variation in conceptualization of similarity across cultural groups,
variation in the input to similarity judgment (namely, environmental
statistics as proxied by language statistics) can be used to predict
cross-cultural differences observed in how people evaluate similarity.
We note that our study is correlational and cannot make a causal claim
about the relationship of environmental statistics to similarity
judgment, but it may be possible to gain traction on potential
mechanisms by examining whether variation in similarity judgments covary
with environmental statistics that differ across cultural and linguistic
contexts. Our specific research questions are as follows:~

Q1: Do we replicate Ji et al. (2004) and extend the results to another
East Asian culture, namely Vietnam? Vietnam is a Southeast Asian country
that borders China and is historically greatly influenced by Chinese
culture (Hui, 2002). Therefore, it serves as a suitable cultural context
to investigate whether the claim made by Ji et al. (2004) and previous
studies, that Eastern and Western cultures have different notions of
similarity, extends beyond mainland China.

Q2: To what extent do cross-context environmental statistics (as proxied
by language) align with variation in similarity judgment?

@@

Q4: Is language statistics viable as a standalone predictor, or is it
simply measuring conceptualization of similarity in a different way?~

To preview our results, we find that, while we do not find the predicted
Eastern/Western contrast in conceptualization of similarity in our
Vietnam extension, we do find that both language-specific statistics and
cultural contexts are a good predictor for cross-cultural differences in
similarity judgments. Language statistics are also a good predictor for
filler items with no taxonomic/thematic structure. Our findings are in
contrast to previous accounts, by which culture induces differing
conceptions of similarity (varying between taxonomic and thematic).
these findings provide an alternative, and more specific, account by
which language may explain these cross-cultural differences without
invoking variation in notions of similarity.

@@

{\centering \includegraphics{figs/unnamed-chunk-4-1}

\caption[Proportion of Word1 responses by country]{Proportion of Word1 responses by country.}\label{fig:unnamed-chunk-4}
\end{figure}
\end{CodeChunk}

For each filler triad, we randomly assigned one of the responding
options as `Word1'. Running a mixed-effects logistic regression to
predict responding (Word1 or Word2) with country as the fixed effect and
a random effect structure equivalent to Q1, we found no effect of
country on filler responding (\(\chi^2\)(2) = 0.74, p = 0.692) (Figure
3).

@@

In this paper, we consider whether statistics of the environment (as
proxied by langauge statistics) can account for cross-cultural
differences in a classic similarity judgment paradigm, as an alternative
to the view that members of different cultures vary in their
conceptualization of similarity. While we managed to replicate the
previously documented contrast between English speakers in the US and
Mandarin Chinese speakers from East Asia (mainland China, Taiwan, Hong
Kong, and Singapore), we do not extend this contrast to our sample of
Vietnamese speakers in Vietnam and English speakers in the US. This
finding suggests some limitations on the generality of the cultural
account. We found some support for the environmental statistics account:
each corpus statistics is a good predictor for the corresponding
country's similarity judgments, even when other corpus statistics are
included, and even with triads without a taxonomic/thematic structure.
However, we also found that culture-based conceptualization of
similarity uniquely contributes to predicting similarity judgment when
included in a model with corpus statistics. Overall, our results provide
evidence that cross-cultural differences in similarity judgment might
not be driven by cross-cultural differences in conception of similarity
alone, but also by variation in inputs to similarity judgment.

@@

Place acknowledgments (including funding information) in a section at
the end of the paper.

>>>>

\begin{abstract}
Is ``cow'' more closely related to ``grass'' or ``chicken''? Speakers of
different languages judge similarity in this context differently, but
why? One possibility is that cultures co-varying with these languages
induce differences in conceptualizations of similarity. Specifically,
East Asian cultures may promote reasoning about thematic similarity, by
which cow and grass are more related, whereas Western cultures may bias
judgments toward taxonomic relations, like cow-chicken. This difference
in notions of similarity is the consensus interpretation for
cross-cultural variation in this paradigm. We consider, and provide
evidence for, an alternative possibility, by which notions of similarity
are similar across contexts, but the statistics of the environment vary.
On this account, similarity judgments are guided by co-occurrence in
experience, and hearing about cows and grass or cows and chickens more
often could induce preferences for these groupings, and account in part
for apparent differences in notions of similarity across contexts.

@@

We constantly partition our continuous experience of the world into
categories, but the way we organize these partitions can vary across
cultures. One way this cross-cultural difference has been observed is
with a task comparing preference for taxonomic and thematic similarity
between East Asian and Western participants.\footnote{Taxonomic
  categorization is based on the similarity of attributes, for example,
  similar perceptual properties, like shared color or shape, among
  objects.In contrast, thematic categorization is based on causal,
  spatial, and temporal relationships among objects (Markman \&
  Hutchinson, 1984).} Ji, Zhang, \& Nisbett (2004) found that in a word
triad task (choose two out of three words that are more related to one
another), Chinese participants preferred thematic matching more compared
to European Americans. In a pictorial version of this task, Chiu (1972)
found that Chinese children (9-10 years old) are also more likely to
choose thematic matches compared to their American counterparts. This
cross-cultural difference is also observed in novel object
categorization, with Chinese participants preferring to group by family
resemblance across multiple features and Americans preferring a
single-feature rule (Norenzayan, Smith, Kim, \& Nisbett, 2002).

One hypothesis is that differences in similarity judgment is linked to
the tendency to see more analytic processing in Western culture and more
holistic processing in East Asian cultures (Nisbett (2003)). On this
hypothesis, since analytic processing emphasizes relationships between
objects and their properties, it leads to taxonomic similarity
judgments; in contrast, holistic processing emphasizes relations between
objects and leads to thematic similarity judgments. Cross-cultural
difference in similarity judgment is thus connected to work showing East
Asian participants exhibiting a higher level of sensitivity to context
than their Western counterparts when reproducing drawings from memory
(Ji, Peng, \& Nisbett, 2000); visually exploring naturalistic scenes
(Chua, Boland, \& Nisbett, 2005); describing scenes (Masuda \& Nisbett,
2001); and in explaining the causes of ambiguous behaviors (Choi,
Nisbett, \& Norenzayan, 1999). This view ascribes cross-cultural
differences in similarity judgment to variation in how each culture (or
group of culture, i.e.~East Asian vs Western) \emph{conceptualize}
similarity -- with East Asians preferring thematic relations more than
Westerners.

Alternatively, these judgments could be shaped by cross-cultural
differences in the \emph{inputs} to similarity judgment, namely, the
statistics of the environment, and accordingly the content of everyday
experiences. Perhaps when faced with the triad task, participants from
all cultures follow the same process for conceptualizing similarity, but
rely on language or culture-specific inputs to this process. If we
observe a difference in categorization between East Asian and Western
participants, it could be that members of both groups use the same
procedure (considering similarity that is influenced by both taxonomic
and thematic relations), but the inputs to this procedure differ between
cultures, with East Asian participants exposed to more support for
thematic similarity in the language they hear in comparison to their
Western counterparts. We might also expect that both conceptualization
and inputs of similarity play a role in driving cross-cultural
differences -- East Asian participants may both be exposed to more
instances of thematic similarity, and prefer to conceptualize similarity
as thematic relations more so than Westerners.

@@

To test alternative hypotheses involving variation of inputs to
similarity judgment, we need a way to operationalize variation in
exposure to thematic vs taxonomic relations. While this metric is
difficult to directly measure, language statistics can provide a rough
proxy. Language statistics is useful in that it is part of the inputs to
everyday experiences (through spoken and written language), may afford
many of the `experiences' that people have with infrequently encountered
items, like cows or helicopters, and provides accessible measures.
Previous work also suggests that using language statistics (such as
lexical co-occurrence or cosine distance of word embeddings) as a proxy
can be a good predictor for similarity reasoning. A model trained on
word-document co-occurrence can predict word association and the effects
of semantic association on a variety of linguistic tasks (Griffiths,
Steyvers, \& Tenenbaum, 2007). Lexical semantic models that are
constructed using lexical co-occurrence (in comparison to annotated
relations) have been shown to perform well on predicting human judgments
about similarity between word pairs that are thematically or
taxonomically related (Rohde, Gonnerman, \& Plaut, 2006). Word
embeddings like word2vec, gloVe, and fastText have been demonstrated to
be good predictors for similarity judgments (Liu, Feng, Wu, Chan, \&
Fulton (2019), Jatnika, Bijaksana, \& Suryani (2019)).

@@

Our study tests the hypothesis that, rather than (or in addition to) the
variation in conceptualization of similarity across cultural groups,
variation in the input to similarity judgment (namely, environmental
statistics as proxied by language statistics) can be used to predict
cross-cultural differences observed in how people evaluate similarity.

We measured taxonomic versus thematic categorization in a word triad
task in three populations. Following Ji et al. (2004), we conducted our
study in the US and China. In addition, we collected data in Vietnam,
Vietnam is a Southeast Asian country that borders China and is
historically greatly influenced by Chinese culture (Hui, 2002).
Therefore, it serves as a suitable cultural context to investigate
whether the claim made by Ji et al. (2004) and previous studies, that
Eastern and Western cultures have different notions of similarity,
extends beyond mainland China. We then tested if fastText vectors from
the corresponding corpus (English, Mandarin, and Vietnamese) are good
predictors for similarity judgment in each cultural context.

We note that our study is correlational and cannot make a causal claim
about the relationship of environmental statistics to similarity
judgment, but it may be possible to gain traction on potential
mechanisms by examining whether variation in similarity judgments covary
with environmental statistics that differ across cultural and linguistic
contexts. Our specific research questions are as follows:~

Q1: Do we replicate cross-cultural differences in thematic/taxonomic
categorization between East Asian and Western cultures?

Q2: Are these cross-cultural differences related to differences in
lexical co-occurrence?

@@

Q4: Is the relationship with co-occurrence specific to
taxonomic/thematic items, or can co-occurrence structure predict other
judgments?~

To preview our results, we replicate US-China differences and find that
Vietnamese judgments are intermediate between these two. Further,
language-specific statistics are a good predictor for cross-cultural
differences in similarity judgments. Language statistics are also a good
predictor for filler items with no taxonomic/thematic structure. Our
findings are in contrast to previous accounts, by which culture induces
differing conceptions of similarity (varying between taxonomic and
thematic). these findings provide an alternative account by which
language may explain these cross-cultural differences without invoking
variation in notions of similarity.

@@

For each filler triad, we randomly assigned one of the responding
options as `Word1'. Running a mixed-effects logistic regression to
predict responding (Word1 or Word2) with country as the fixed effect and
a random effect structure equivalent to Q1, we found no effect of
country on filler responding (\(\chi^2\)(2) = 0.74, p = 0.692).

@@

In this paper, we consider whether statistics of the environment (as
proxied by language statistics) can account for cross-cultural
differences in a classic similarity judgment paradigm, as an alternative
to the view that members of different cultures vary in their
conceptualization of similarity. While we managed to replicate the
previously documented contrast between English speakers in the US and
Mandarin Chinese speakers from East Asia (mainland China, Taiwan, Hong
Kong, and Singapore), we do not extend this contrast to our sample of
Vietnamese speakers in Vietnam and English speakers in the US. This
finding suggests some limitations on the generality of the cultural
account. We found some support for the environmental statistics account:
each corpus statistics is a good predictor for the corresponding
country's similarity judgments, even when other corpus statistics are
included, and even with triads without a taxonomic/thematic structure.
However, we also found that culture-based conceptualization of
similarity uniquely contributes to predicting similarity judgment when
included in a model with corpus statistics. Overall, our results provide
evidence that cross-cultural differences in similarity judgment are
related to linguistic co-occurrence patterns (which also vary across
cultures).

@@

BLINDED FOR PEER REVIEW.
